% !TEX ROOT = ../Thesis.tex 
%-
%-> 中文摘要
%-
\chapter{摘\quad 要}\chaptermark{摘\quad 要}% 摘要标题
% \setcounter{page}{1}% 开始页码
% \pagenumbering{Roman}% 页码符号
% 22p / 12p = 1.83
\linespread{1.5}
\zihao{-4}
% \setlength{\baselineskip}{20pt}

在大型语言模型与检索增强生成技术飞速发展的背景下，多轮对话检索系统的召回准确性已成为决定人工智能在复杂工业场景中落地能力的核心要素。然而，现有多轮对话检索基准数据集普遍面临三大严峻挑战：人工标注成本高昂且扩展性差；人类固有“认知边界”导致严重的标注稀疏性；以及传统宏观评估指标在复杂语义对齐与幻觉检测上存在系统性盲区。这些缺陷导致现有测试基准呈现严重的“分数饱和”现象，无法真实反映现代先进检索系统的能力边界。针对上述痛点，本研究提出了一套涵盖细粒度质量评估、自动化高保真合成与极限压力评测的多轮对话检索完整解决方案。本文的主要研究工作与创新成果如下：

首先，确立了突破人类认知边界的细粒度自动化审计体系MTR-EVAL。针对传统信息检索指标难以量化大模型生成质量的问题，本研究基于“大语言模型作为裁判（LLM-as-a-judge）”范式，将数据质量解构为查询-证据对齐度、证据完备性、答案-证据忠实度及答案质量四个正交维度。通过引入鉴别性测试与多模型集成的去偏机制，该体系在实证检验中实现了与人类专家打分的高度正相关（答案-证据忠实度皮尔逊相关系数高达0.8924），彻底解决了基准数据的质量审计难题，为后续的数据合成提供了坚实的量化度量衡。

其次，构建了基于图论规划与多智能体协同的全自动数据合成流水线MTR-Pipeline。为突破高质量数据构建的成本与规模化壁垒，在底层话题规划阶段，本研究创新性地提出了基于旅行商问题贪婪遍历的文档语义轨迹聚类算法，有效消除了数据冗余，并在语义空间中构建了平滑的话题演进网络。在对话生成阶段，设计了由提问者、回答者与润色者构成的三智能体协同架构。通过严格的角色解耦与语言学后处理，该流水线隐式注入了真实的指代与省略等口语化特征，并将合成单段高质量多轮对话的成本大幅降低至传统人工众包标注的1/400。

最后，提出了极具区分度的工业级评测基准MTR-BENCH，并深度揭示了现有检索系统的技术盲区。依托自动化流水线，本研究以最新维基百科（2025版）为知识库，构建了包含近4万个交互轮次的大规模评测基准，显式注入了软/硬话题双重切换与工业级长文本冗余等高难度特征。大规模基线评估实验表明，当前业界顶尖的单轮与多轮检索模型在面对MTR-BENCH时，其Recall@20核心指标较传统数据集出现了超过43个百分点的断崖式下降。结合大语言模型先知重写实验与金融垂直领域的泛化消融实验，本研究确凿地证实了该性能落差来源于现代检索模型在深层多轮语义解耦与抗噪机制上的技术盲区，而非底层标注噪声。

综上所述，本研究不仅在理论层面完成了多轮对话基准从“局部启发式标注”向“全局感知自动化合成”的范式跃迁，更为学术界和工业界研发与迭代新一代会话检索系统，提供了一套高度纯净、极具区分度测试环境。

\keywords{多轮对话检索；检索增强生成；大语言模型；自动化数据合成}% 中文关键词

%-
%-> 英文摘要
%-
\chapter{Abstract}\chaptermark{Abstract}% 摘要标题

In the context of the rapid development of large language models (LLMs) and retrieval-augmented generation (RAG) technologies, the recall accuracy of multi-turn conversational retrieval systems has become a core factor determining the practical deployment capability of artificial intelligence in complex industrial scenarios. However, existing multi-turn conversational retrieval benchmark datasets generally face three severe challenges: high cost and poor scalability of manual annotation; inherent “cognitive boundaries” of humans leading to severe annotation sparsity; and systematic blind spots of traditional macro-level evaluation metrics in handling complex semantic alignment and hallucination detection. These deficiencies result in a pronounced “score saturation” phenomenon in existing benchmarks, making them incapable of truly reflecting the capability boundaries of modern advanced retrieval systems. To address these issues, this study proposes a comprehensive solution for multi-turn conversational retrieval, encompassing fine-grained quality evaluation, automated high-fidelity data synthesis, and extreme stress testing. The main research contributions and innovations of this work are as follows:

First, we establish a fine-grained automated auditing framework, MTR-EVAL, to transcend human cognitive boundaries. To address the difficulty of quantifying generation quality using traditional information retrieval metrics, this study adopts the “LLM-as-a-judge” paradigm and decomposes data quality into four orthogonal dimensions: query–evidence alignment, evidence completeness, answer–evidence faithfulness, and answer quality. By introducing discriminability testing and a debiasing mechanism based on multi-model integration, the framework achieves a high positive correlation with human expert judgments in empirical evaluations (with a Pearson correlation coefficient of up to 0.8924 for answer–evidence faithfulness), effectively solving the problem of quality auditing for benchmark data and providing a solid quantitative foundation for subsequent data synthesis.

Second, we construct a fully automated data synthesis pipeline, MTR-Pipeline, based on graph-theoretic planning and multi-agent collaboration. To overcome the cost and scalability barriers in high-quality data construction, at the underlying topic planning stage, this study innovatively proposes a document semantic trajectory clustering algorithm based on greedy traversal of the Traveling Salesman Problem, effectively eliminating data redundancy and constructing a smooth topic evolution network in the semantic space. In the dialogue generation stage, a three-agent collaborative architecture consisting of a questioner, an answerer, and a refiner is designed. Through strict role decoupling and linguistic post-processing, the pipeline implicitly injects natural conversational features such as coreference and ellipsis, while significantly reducing the cost of synthesizing a single high-quality multi-turn dialogue to 1/400 of traditional human crowd-sourced annotation.

Finally, we produce a highly discriminative industrial-grade evaluation benchmark, MTR-BENCH, and deeply reveal the technical blind spots of existing retrieval systems. Leveraging the automated pipeline, this study constructs a large-scale benchmark based on the latest Wikipedia (January 2025 dump), containing nearly 40,000 interaction turns, with explicitly injected challenging features such as soft/hard topic shifts and industrial-scale long-text redundancy. Large-scale baseline experiments demonstrate that state-of-the-art single-turn and multi-turn retrieval models experience a dramatic drop of over 43 percentage points in the core Recall@20 metric when evaluated on MTR-BENCH compared to traditional datasets. Combined with LLM oracle rewriting experiments and generalization ablation studies in the financial domain, this study conclusively verifies that the performance gap originates from technical blind spots in deep multi-turn semantic disentanglement and noise robustness of modern retrieval models, rather than underlying annotation noise.

In summary, this study not only achieves a paradigm shift in multi-turn conversational benchmarks from “local heuristic annotation” to “global perception-driven automated synthesis” at the theoretical level, but also provides a highly clean, strongly discriminative infrastructure and testing environment for the development and iteration of next-generation conversational retrieval systems in both academia and industry.


\englishkeywords{multi-turn conversational retrieval; retrieval-augmented generation; large language models; automated data synthesis}% 


